---
title: "data.table: fast and efficient extension of data.frame"
output: ioslides_presentation
---

```{r include=F}
library(microbenchmark)
library(sqldf)
library(dplyr)
library(data.table)
library(formattable)

wd <- 'H:/scripts/r/datatable_presentation'

options(datatable.print.topn=3)
options(datatable.print.nrows=5)

set.seed(42)
```

## motivations
* `data.table` is significantly faster and more memory efficient than any other R options  
* some performance comparisons for left joining two datasets of 50M rows ([source](https://stackoverflow.com/questions/1299871/how-to-join-merge-data-frames-inner-outer-left-right/34219998#34219998)):  

```{r echo=F}
# n = 5e7
# set.seed(108)
# df1 = data.frame(x=sample(n,n-1L), y1=rnorm(n-1L))
# df2 = data.frame(x=sample(n,n-1L), y2=rnorm(n-1L))
# dt1 = as.data.table(df1)
# dt2 = as.data.table(df2)
# 
# mb <- microbenchmark(
#   times = 1L,
#   base = merge(df1, df2, by = "x", all.x = TRUE),
#   sqldf = sqldf("SELECT * FROM df1 LEFT OUTER JOIN df2 ON df1.x = df2.x"),
#   dplyr = left_join(df1, df2, by = c("x"="x")),
#   DT = dt2[dt1, on = "x"]
# )
# dtmb <- as.data.table(summary(mb))
# fwrite(dtmb, file = file.path(wd, 'dtmb.csv'))
dtmb <- fread(file = file.path(wd, 'dtmb.csv'))
dtmb[order(-min), .(Method = expr, Time = comma(min, digits = 2))]
```

# basics
## construction
```{r}
indicators <- 1:100
x <- data.table(
  indicator = rep(indicators, each = 26),
  group = rep(letters, times = length(indicators)),
  measurement = rnorm(26 * length(indicators))
)
x[]
```

## syntax | x[i, j, by]

# i
## basic subsetting
```{r}
x[2, ]
x[2]
```

## predicate subsetting
```{r}
x[indicator %in% c(1:3, 10:20), ]
```

## predicate subsetting
```{r}
x[group %like% '[ab]', ] # %like% provided by data.table
```

# j
## j
```{r}
x[, .(group, measurement)]
```

# by
## by
```{r}
x[, .(group, mean(measurement))]
```

## by
```{r}
x[, .(mean(measurement)), by = .(group)]
```

# Special Symbols
## .N
```{r}
x[, .N, by = .(group)]
```

## .N
```{r}
x[.N, ]
x[, .N]
```

## .SD/.SDcols
```{r}
x[
  ,
  .SD,
  .SDcols = c('group', 'measurement')
]
```

## .SD/.SDcols
```{r}
x <- cbind(x, measurement2 = runif(x[, .N]))

x[
  ,
  lapply(.SD, mean),
  by = .(indicator),
  .SDcols = c('measurement', 'measurement2')
]
```

## .I
```{r}
grpcols <- c('group')
x[
  ,
  head(.SD, 1),
  by = grpcols,
  .SDcols = c('measurement')
]
```

## .I
```{r}
x[
  x[
    ,
    .I[abs(measurement - max(measurement)) < sqrt(.Machine$double.eps)],
    by = grpcols
  ]$V1,
]
```

## :=
```{r}
x[2, ]
x[2, measurement := rnorm(1)]
x[2, ]
```

## :=
```{r}
x[
  ,
  measurement3 := mean(measurement) - mean(measurement2),
  by = .(indicator)
]
x[]
```

## :=
```{r}
x[, c(paste0('measurement', 2:3)) := NULL]
x[]
```

# on
## standard SQL joins
### right[left, on = joincols, nomatch = c(NA, NULL)]
## standard SQL joins
```{r}
observers <- rbind(
  data.table(observer = 'nick', group = head(letters, 3)),
  data.table(observer = 'schmidt', group = tail(letters, 3))
)
observers[]
```

## standard SQL joins | left join
```{r}
observers[x, on = c('group')]
```

## standard SQL joins | right join
```{r}
x[observers, on = c('group')]
```

## rolling joins
```{r}
tools <- data.table(
  indicator = c(1, 51),
  tool = c('iphone', 'samsung galaxy')
)
tools[]
```

## rolling joins
```{r}
tools[x, on = c('indicator'), roll = Inf]
```

## notjoins
```{r}
x[!observers, on = c('group')]
```

## advanced subsetting
```{r}
x[.(14:15), on = c('indicator')]
```

## advanced subsetting | CJ
```{r}
x[
  CJ(indicator = 7:9, group = c('f', 'g', 'y')),
  on = c('indicator', 'group')
]
```

## advanced subsetting | CJ
```{r}
x <- x[!sample(.N, size = 20), ]
x[]
```

## advanced subsetting | CJ
```{r}
x <- x[
  CJ(indicator, group, unique = T),
  on = c('indicator', 'group')
]
x[]
```

## advanced subsetting | CJ
```{r}
x[!complete.cases(x), ]
```

# data.table methods
## fread/fwrite
* `data.table`'s .*sv reading and writing functions  
* like rest of package, extremely fast and memory-efficient  
* highly customizable; I pretty much never do anything except  
      + `fread(file = 'path/to/file.csv')  
      + fwrite(x, file = 'path/to/file.csv')  

## order
```{r}
x[order(-measurement), ]
```

## reshaping | dcast
```{r}
xwide <- dcast(x, indicator ~ group, value.var = 'measurement')
xwide[, .SD, .SDcols = head(names(xwide), 6)]
```

## reshaping | melt
```{r message=F}
melt(xwide, id.vars = c('indicator'))
```

## shift
```{r}
x[, shift(measurement, n = 1:2, type = 'lag'), by = .(group)]
```

## frollapply
* two common uses `frollmean` and `frollsum` are also available
```{r}
x[, frollapply(measurement, n = 3, min), by = .(indicator)]
```

# questions
